---
title: "Simulations In-Class Project"
date: "Due October 13, 2017 at 11:59pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:


With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
```{r}

gamblers<-function(initial,games,bet,p){
 
counter=0
a<-(rep(0,5000))# this counts the number of games that went bankrupt on the 100th bet

b<-matrix(rep(NA),nrow=1000,ncol=5001)

for ( i in 1:1000 ){       # this is the rows
  
x<-rbinom(5000,1,p) # let p be the probaility of loosing
b[i,1]=1000
current=1000
for ( j in 1:5000){ #inner is the columns
    

    if(x[j]==1 & current>0){  # a value of 1 means I won the bet
      
      current<-current+(100)  ## TRY CHANGING THE INDEX TOMORROW
       b[i,j+1]=current
       
    }else if  (x[j]==0 & current>0){
    current<-current-(100)
    b[i,j+1]=current
   } else if (current<= 0 ){
      b[i,(j+1):5001]=0
      a[j]=1
      
     break
      
    }
  }
   # Count if you go bankrupt at the 100th bet
  
  
}
for ( i in 1:nrow(b)){
  if(b[i,bet]==0 ){
    counter=counter+1
  }
  answer<-counter/1000
 
}
print(answer)
}
gamblers(1000,1000,100,0.50)
```

    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
```{r}
set.seed(10)
gamblers(1000,1000,500,0.50)

```

    c. the mean time you go bust, given that you go bust within the first 5000 hands. # how many hands on average does it take for you to go bust ( about 500)
```{r}
foo<-function(b){

 which(b <= 0)[1]
}

counter=0
a<-(rep(0,5000))# this counts the number of games that went bankrupt on the 100th bet

b<-matrix(rep(NA),nrow=1000,ncol=5001)

for ( i in 1:1000 ){       # this is the rows
  
x<-rbinom(5000,1,0.50) # let p be the probaility of loosing
b[i,1]=1000
current=1000
for ( j in 1:5000){ #inner is the columns
    

    if(x[j]==1 & current>0){  # a value of 1 means I won the bet
      
      current<-current+(100)  ## TRY CHANGING THE INDEX TOMORROW
       b[i,j+1]=current
       
    }else if  (x[j]==0 & current>0){
    current<-current-(100)
    b[i,j+1]=current
   } else if (current<= 0 ){
      b[i,(j+1):5001]=0
      a[j]=1
      
     break
      
    }
  }
   # Count if you go bankrupt at the 100th bet
    vects<-apply(b,1,foo)
mean(vects,na.rm=TRUE)
  
}

```


    d. the mean and variance of your bankroll after 100 hands (including busts).
```{r}
counter=0
a<-(rep(0,5000))# this counts the number of games that went bankrupt on the 100th bet

b<-matrix(rep(NA),nrow=1000,ncol=5001)

for ( i in 1:1000 ){       # this is the rows
  
x<-rbinom(100,1,0.50) # let p be the probaility of loosing
b[i,1]=1000
current=1000
for ( j in 1:100){ #inner is the columns
    

    if(x[j]==1 & current>0){  # a value of 1 means I won the bet
      
      current<-current+(100)  ## TRY CHANGING THE INDEX TOMORROW
       b[i,j+1]=current
       
    }else if  (x[j]==0 & current>0){
    current<-current-(100)
    b[i,j+1]=current
   } else if (current<= 0 ){
      b[i,(j+1):5001]=0
      a[j]=1
      
     break
      
    }
  }
   # Count if you go bankrupt at the 100th bet
    vects<-apply(b,1,mean,na.rm=TRUE)
    mean(vects)
    variance<-apply(b,1,var,na.rm=TRUE)
    mean(variance)
}
```

    e. the mean and variance of your bankroll after 500 hands (including busts).
 
```{r}
counter=0
a<-(rep(0,5000))# this counts the number of games that went bankrupt on the 100th bet

b<-matrix(rep(NA),nrow=1000,ncol=5001)

for ( i in 1:1000 ){       # this is the rows
  
x<-rbinom(500,1,0.50) # let p be the probaility of loosing
b[i,1]=1000
current=1000
for ( j in 1:500){ #inner is the columns
    

    if(x[j]==1 & current>0){  # a value of 1 means I won the bet
      
      current<-current+(100)  ## TRY CHANGING THE INDEX TOMORROW
       b[i,j+1]=current
       
    }else if  (x[j]==0 & current>0){
    current<-current-(100)
    b[i,j+1]=current
   } else if (current<= 0 ){
      b[i,(j+1):5001]=0
      a[j]=1
      
     break
      
    }
  }
   # Count if you go bankrupt at the 100th bet
    vects<-apply(b,1,mean,na.rm=TRUE)
    mean(vects)
    variance<-apply(b,1,var,na.rm=TRUE)
    mean(variance)
}
```

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.
```{r}
gamblers(1000,1000,100,18/38)
```
b
```{r}
gamblers(1000,1000,500,18/38)
```
c.
```{r}
foo<-function(b){

 which(b <= 0)[1]
}

counter=0
a<-(rep(0,5000))# this counts the number of games that went bankrupt on the 100th bet

b<-matrix(rep(NA),nrow=1000,ncol=5001)

for ( i in 1:1000 ){       # this is the rows
  
x<-rbinom(5000,1,18/38) # let p be the probaility of loosing
b[i,1]=1000
current=1000
for ( j in 1:5000){ #inner is the columns
    

    if(x[j]==1 & current>0){  # a value of 1 means I won the bet
      
      current<-current+(100)  ## TRY CHANGING THE INDEX TOMORROW
       b[i,j+1]=current
       
    }else if  (x[j]==0 & current>0){
    current<-current-(100)
    b[i,j+1]=current
   } else if (current<= 0 ){
      b[i,(j+1):5001]=0
      a[j]=1
      
     break
      
    }
  }
   # Count if you go bankrupt at the 100th bet
    vects<-apply(b,1,foo)
mean(vects,na.rm=TRUE)
  
}
```

d.
```{r}
counter=0
a<-(rep(0,5000))# this counts the number of games that went bankrupt on the 100th bet

b<-matrix(rep(NA),nrow=1000,ncol=5001)

for ( i in 1:1000 ){       # this is the rows
  
x<-rbinom(100,1,18/38) # let p be the probaility of loosing
b[i,1]=1000
current=1000
for ( j in 1:100){ #inner is the columns
    

    if(x[j]==1 & current>0){  # a value of 1 means I won the bet
      
      current<-current+(100)  ## TRY CHANGING THE INDEX TOMORROW
       b[i,j+1]=current
       
    }else if  (x[j]==0 & current>0){
    current<-current-(100)
    b[i,j+1]=current
   } else if (current<= 0 ){
      b[i,(j+1):5001]=0
      a[j]=1
      
     break
      
    }
  }
   # Count if you go bankrupt at the 100th bet
    vects<-apply(b,1,mean,na.rm=TRUE)
    mean(vects)
    variance<-apply(b,1,var,na.rm=TRUE)
    mean(variance)
}
```


e.
```{r}
counter=0
a<-(rep(0,5000))# this counts the number of games that went bankrupt on the 100th bet

b<-matrix(rep(NA),nrow=1000,ncol=5001)

for ( i in 1:1000 ){       # this is the rows
  
x<-rbinom(500,1,18/38) # let p be the probaility of loosing
b[i,1]=1000
current=1000
for ( j in 1:500){ #inner is the columns
    

    if(x[j]==1 & current>0){  # a value of 1 means I won the bet
      
      current<-current+(100)  ## TRY CHANGING THE INDEX TOMORROW
       b[i,j+1]=current
       
    }else if  (x[j]==0 & current>0){
    current<-current-(100)
    b[i,j+1]=current
   } else if (current<= 0 ){
      b[i,(j+1):5001]=0
      a[j]=1
      
     break
      
    }
  }
   # Count if you go bankrupt at the 100th bet
    vects<-apply(b,1,mean,na.rm=TRUE)
    
    variance<-apply(b,1,var,na.rm=TRUE)
   
}
average<-mean(vects)
variance<-mean(variance)
```

3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
```{r}
Markov<-function(start,starting_prob, wager,inc,bets){
  
  p<-rep(NA,bets+1)
  bankroll<-rep(NA,bets+1)
  bankroll[1]<-start
  p[1]<-starting_prob
  
for ( i in 1:bets){
  if(rbinom(1,1,p[i])==1 ){ # if you win 
    bankroll[i+1]<-bankroll[i]+wager
    if(p[i]+inc <=1){ #if the current probability plus the increment is less than or equal to 1
      p[i+1]<-p[i]+inc # update the new probability by adding an increment
    }
  else{
    1<-p[i+1] # keep the probability the same
  }
  }else{  #if you loose 
    bankroll[i+1]<-bankroll[i]-wager
    starting_prob=p[i+1]
  } 
 
    
}

  
return(bankroll[i])  # return the final amount of money i have
}

```

```{r}
Markov<-function(p){
return <- numeric(100)
for(trial in 1:100){
  hands <- numeric(100000)
  #p <- .48
  hands[1] <- 1
  for(i in 2:100000){
    if(hands[i-1] == 1 & p <= .98){
      p <- p + .02
    } else {
      p <- .48
    }
    hands[i] <- rbinom(1, 1, p)
  }
  return[trial] <- mean(hands)
}
mean(return)
}
```

    
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
```{r}
Markov(0.46)
Markov(0.47)
Markov(0.472)
```

    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
```{r}
return <- numeric(10000)
for(trial in 1:10000){
  hands <- numeric(100)
  p <- .48
  hands[1] <- 1
  for(i in 2:100){
    if(hands[i-1] == 1 & p <= .98){
      p <- p + .01
    } else {
      p <- .48
    }
    hands[i] <- rbinom(1, 1, p)
  }
  return[trial] <- mean(hands)
}
mean(return)
```
```{r}
return <- numeric(10000)
for(trial in 1:10000){
  hands <- numeric(100)
  p <- .48
  hands[1] <- 1
  for(i in 2:100){
    if(hands[i-1] == 1 & p <= .98){
      p <- p + .015
    } else {
      p <- .48
    }
    hands[i] <- rbinom(1, 1, p)
  }
  return[trial] <- mean(hands)
}
mean(return)
```





4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile

Make a function called `boot_ci` which calculates the 95% confidence interval in this manner. 
```{r}
boot_ci<-function(data,itteration,statistic){
  stats<-rep(NA,length=itteration)
  for ( i in 1:itteration){
    

samples<-sample(data,size=length(data),replace=TRUE)
stats[i]<-statistic(samples)
  }


lower<-quantile(stats,0.025)
upper<-quantile(stats,0.975)
CI<-upper-lower

}
```

5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?
```{r}
data<-replicate(100,Markov(0.48))
boot_ci(data,1000,var)
```
 
